---
title: "Response to Reviewers"
subtitle: |
  NPH-MS-2026-57029\
  *Guard cell size and pore aperture influence stomatal closure kinetics*\
  Muir & Lim
format:
  pdf:
    toc: false
    number-sections: false
    indent: false
    fontsize: 11pt
    geometry: margin=1in
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
header-includes: |
  \usepackage{xcolor}
  \usepackage{mdframed}
  \usepackage{setspace}
  \usepackage{placeins}
  \usepackage{siunitx}
  \renewcommand{\thefigure}{R\arabic{figure}}
  \definecolor{responseblue}{RGB}{70, 130, 180}
  \definecolor{commentgray}{RGB}{90, 90, 90}
  \newenvironment{response}
    {\color{responseblue}\setlength{\parskip}{0.5em}}
    {}
  \newenvironment{revcomment}
    {\color{commentgray}\itshape\begin{mdframed}[linecolor=commentgray,linewidth=0.5pt,innerleftmargin=8pt,innerrightmargin=8pt,innertopmargin=6pt,innerbottommargin=6pt]}
    {\end{mdframed}}
  \newcommand{\fgmax}{$f_\mathrm{gmax}$}
  \newcommand{\gcl}{$l_\mathrm{gc}$}
  \newcommand{\gi}{$g_\mathrm{i}$}
  \newcommand{\gmax}{$g_\mathrm{max}$}
  \newcommand{\gop}{$g_\mathrm{op}$}
  \newcommand{\gsw}{$g_\mathrm{sw}$}
---

We thank the editor and both reviewers for their thoughtful and constructive
comments. We address each point below. Reviewer comments appear in gray boxes;
our responses appear in blue.

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(glue)

comp_pars_vpd = read_rds("../objects/tbl-comparison-vpd.rds")
final_vpd_effect = read_rds("../objects/tbl-final-vpd-effect.rds")
cor_tau_rate = read_rds("../objects/cor-tau-vpd-rate.rds")

fmt_est = function(.df, .parameter, .model) {
  row = .df[grepl(.parameter, .df$parameter, fixed = TRUE) & .df$model == .model, ]
  glue(
    "{formatC(row$estimate, format = 'f', digits = 2)} ",
    "[{formatC(row$lower, format = 'f', digits = 2)}, ",
    "{formatC(row$upper, format = 'f', digits = 2)}]"
  )
}

fgmax_orig = fmt_est(comp_pars_vpd, "f[gmax]", "Original")
fgmax_vpd = fmt_est(comp_pars_vpd, "f[gmax]", "VPD-adjusted")
gcl_orig = fmt_est(comp_pars_vpd, "phylogenetic~corr", "Original")
gcl_vpd = fmt_est(comp_pars_vpd, "phylogenetic~corr", "VPD-adjusted")

null_sim_results = read_rds("../objects/null-sim-fgmax-tau.rds")
real_cor_fgmax_tau = read_rds("../objects/real-cor-fgmax-tau.rds")
null_sim_p = read_rds("../objects/null-sim-empirical-p.rds")
nls_vs_bayes = read_rds("../objects/null-sim-nls-vs-bayes-summary.rds")

fmt_cor = function(x) formatC(x, format = "f", digits = 2)

final_vpd_on_tau = glue(
  "{formatC(final_vpd_effect$estimate, format = 'f', digits = 2)} ",
  "[{formatC(final_vpd_effect$lower, format = 'f', digits = 2)}, ",
  "{formatC(final_vpd_effect$upper, format = 'f', digits = 2)}]"
)
```

---

# Reviewer 1

## Comment R1.1 — Realized VPD as a covariate

::: revcomment
It seems like the realized VPD stimulus reported varies systematically among
treatments and this might needs to be treated as part of the analysis rather
than as a supplementary caveat, as in Lines 238–249 are unusually candid and
useful, as the authors state that scrubbing incoming water vapor "did not
standardize chamber air humidity because leaf transpiration introduced water
vapor," and that RH was systematically higher at high measurement light, in
sun grown plants and in amphistomatous measurements because \gsw{} was greater
in those conditions. Table S3 then shows that the mean VPD during measurements
ranged from 2.50 kPa in sun/high/amphi curves to 2.92 kPa in
shade/low/pseudohypo curves. I think this matters because the actual forcing
variable in the experiment is not simply the experimental treatment label, but
also is the VPD trajectory experienced by the leaf. If leaves with higher
conductance both have higher \fgmax{} and also humidify the chamber more
strongly, then \fgmax{} is partly entangled with the realized stimulus. I would
love to see the authors include curve level VPD covariates in the model such as
median VPD during the fitted interval, final VPD if available, VPD step
magnitude, and perhaps the rate at which VPD changed. At minimum, should show
whether the \fgmax{} → $\tau$ effect remains after accounting for actual VPD
exposure, and ideally they should run a matched VPD sensitivity analysis
excluding curves outside a narrow VPD range. I would be careful describing the
VPD treatment as though the desiccant step created a common perturbation across
treatments unless this is demonstrated. This distinction is important because
VPD is a primary driver of stomatal conductance and plant water relations, and
the physiologically relevant stimulus is the leaf to air evaporative gradient
not simply the inlet humidity setting.
:::

::: response
**Response:**

We added Notes S1 to the Supporting Information to address the potential for variable VPD stimuli to confound our results, and updated the Materials and Methods to point to it. Our main results are robust once variation in VPD is incorporated into the analysis: the fixed effect of \fgmax{} on $\log \tau$ (original: `r fgmax_orig`; VPD-adjusted: `r fgmax_vpd`) and the phylogenetic correlation between \gcl{} and $\tau$ (original: `r gcl_orig`; VPD-adjusted: `r gcl_vpd`) were both essentially unchanged after adding final VPD as a covariate of $\tau$ and $\lambda$ (@fig-vpd-comparison). One caveat is that we do not have data on the initial rate of change in VPD, which limits our ability to evaluate whether this influenced stomatal kinetic responses; we added a recommendation for future research on this point at the end of Notes S1.

We extracted the leaf-to-air VPD trajectory for every curve from the logged humidity data and summarized it with two curve-level covariates: the VPD at the end of the fitted interval ("final VPD") and the rate at which VPD approached that final value ("VPD saturation rate"). We considered adding
median VPD and VPD step magnitude as well, as the reviewer suggested. However, median VPD and final VPD are highly correlated ($r = 0.93$; @fig-vpd-pairs a), so we chose to use the final VPD. This should be the same as the VPD step magnitude since all leaves were started at the same RH and Tleaf, hence starting VPD was approximately constant.

We attempted to incorporate the rate of VPD change as a covariate, but determined this was not useful. Unfortunately, we did not log data during the WWR when the initial step change in VPD occurred. We instead fit a saturating exponential, $\mathrm{VPD}(t) = B + (R_0 - B)\exp(-kt)$, to each curve's VPD trajectory (median $R^2 = 0.99$) and use the rate constant $k$ ("VPD saturation rate"). This rate constant is only weakly correlated with final VPD ($r = 0.31$; @fig-vpd-pairs c). However, $k$ is strongly correlated with $\tau$ ($r = `r formatC(cor_tau_rate, format = "f", digits = 2)`$ on a log-log scale; @fig-tau-vpd-rate) because it is mostly caused by decreasing $g_\mathrm{sw}$. We can see this clearly by comparing the rate of change in the water vapor concentration of the reference airstream, upstream of the leaf ($W_\mathrm{r}$), versus that of the sample airstream, which reflects humidity inside the chamber ($W_\mathrm{s}$). If the change in VPD is caused by $g_\mathrm{sw}$, we expect a positive correlation between the rate of VPD change and the rate of $W_\mathrm{s}$ change, which is exactly what we see (@fig-vpd-h2o-rate a). In contrast, the rate of $W_\mathrm{r}$ change is uncorrelated with the rate of VPD change (@fig-vpd-h2o-rate b). Hence, it would not be statistically valid to include the rate of VPD change as a covariate in predicting $\tau$ since both are caused by the same underlying variable, the rate of change in $g_\mathrm{sw}$.

To directly test whether the realized VPD stimulus confounds the \fgmax{} and \gcl{} effects on $\tau$, we refit our selected model with final VPD added as a covariate of $\tau$ and $\lambda$. As reported above, the \fgmax{} and \gcl{} effects were essentially unchanged (@fig-vpd-comparison). Final VPD itself had a negative fixed effect on $\log \tau$ (`r final_vpd_on_tau`), consistent with higher realized VPD accelerating stomatal closure.

```{r}
#| label: fig-vpd-pairs
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Pairwise relationships among curve-level VPD covariates: (a) median vs. final VPD, (b) median VPD vs. VPD saturation rate, (c) final VPD vs. VPD saturation rate. Each point is one humidity response curve, colored by leaf type; lines are ordinary least-squares fits within leaf type. VPD saturation rate is the rate constant of a saturating exponential fit to the VPD trajectory during the fitted interval."
knitr::include_graphics("../figures/vpd-pairs.pdf")
```

```{r}
#| label: fig-tau-vpd-rate
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The stomatal closure time-constant $\\tau$ is positively correlated with the inverse of the VPD saturation rate, $k^{-1}$ (both log scale; $r = `r formatC(cor_tau_rate, format = 'f', digits = 2)`$), consistent with $k$ being largely driven by the same $g_\\mathrm{sw}$-decline signal that determines $\\tau$, rather than an independent property of the VPD stimulus. Each point is one humidity response curve."
knitr::include_graphics("../figures/tau-vpd-rate.pdf")
```

```{r}
#| label: fig-vpd-h2o-rate
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The VPD saturation rate is driven by the sample airstream, not the reference airstream. (a) VPD saturation rate vs. the saturation rate of $W_\\mathrm{s}$ (sample cell water vapor concentration, which has equilibrated with the leaf); the dashed line is the 1:1 line. (b) VPD saturation rate vs. the saturation rate of $W_\\mathrm{r}$ (reference cell water vapor concentration, upstream of the leaf). Each point is one humidity response curve; rate constants come from the same saturating-exponential model used for VPD."
knitr::include_graphics("../figures/vpd-h2o-rate.pdf")
```

```{r}
#| label: fig-vpd-comparison
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The \\fgmax{} effect on $\\tau$ and the phylogenetic correlation between \\gcl{} and $\\tau$ are robust to accounting for final VPD. Posterior means and 95\\% credible intervals for the fixed effect of \\fgmax{} on $\\log \\tau$ (left) and the phylogenetic correlation between $\\log$ \\gcl{} and $\\log \\tau$ (right), comparing the original model (grey) to a model refit with final VPD added as a covariate of $\\tau$ and $\\lambda$ (pink)."
knitr::include_graphics("../figures/vpd-comparison.pdf")
```
:::

```{=latex}
\FloatBarrier
```

---

## Comment R1.2 — Mislabeled VPD summary statistic

::: revcomment
Related to the previous point, the manuscript appears to call the Table S3
values "final RH" and "final VPD," but I found the code and caption indicate
something different. At lines 245 to 247, the text says that "the final RH"
and "VPD" values in Table S3 elicited rapid stomatal response. However, the
Table S3 caption says the values are calculated as the median RH and VPD
across time points within each curve and then averaged among curves. The code
confirms this: `r/09_make-tbl-vpd.R`, lines 7–14, calculates median RH and
VPD per curve, and lines 32–40 average those medians across treatment
combinations. That is not the same as final RH/VPD nor necessarily the VPD
step magnitude. It would be great if author can correct the wording and, more
importantly, to report the full VPD trajectory more transparently, with
pre-step VPD, VPD at the start of logging, median VPD during the fitted RWR,
final VPD, and ideally the change in VPD over time, because the paper
interprets variation in $\tau$ as intrinsic closure kinetics, while the
stimulus itself may vary across the same treatments used to explain $\tau$.
:::

::: response
**Response:**

We thank the reviewer for catching the discrepancy between what we reported in Table S3 versus the actual content. At noted in the response to the previous comment, the median and final RH and VPD are highly correlated, so this mistake did not affect the interpretation. We revised Table S3 to report the average RH and VPD at three time points: when logging started, when logging finished, and the median during the logging interval. We also added the curve-level values for these variables to Table S1. We did not record the pre-step VPD but it was nearly identical for every leaf because we set leaf temperature to $\SI{25}{\degreeCelsius}$ and 70% to standardize starting conditions. This is equivalent to a VPD of $\mathrm{VPD} \approx \SI{0.82}{\kPa}$, as mentioned in the Materials and Methods. As mentioned in the previous response, we also do not have data on the initial rate of VPD increase.

:::

---

## Comment R1.3 — \fgmax{} is not an independent aperture measurement

::: revcomment
I like \fgmax{} as a quite intriguing predictor but it is not an independent
direct measure of stomatal aperture. The manuscript motivates \fgmax{} as
proportional to aperture relative to maximum aperture at lines 115–133,
defines it as $g_i / g_\mathrm{max}$ at lines 273–279 and then interprets
higher \fgmax{} as stomata being more open at lines 397–404 and as a
mechanistic driver of slower closure at lines 506–513. I think the hypothesis
is interesting, but the current evidence is not enough to treat \fgmax{} as a
direct aperture measurement. The numerator, \gi, is estimated from the same
nonlinear curve that also gives $\tau$ and $\lambda$, so parameter covariance
in the Weibull fit could generate an association between \gi and $\tau$ even
if the underlying biological relationship is weaker. The denominator
$g_\mathrm{max}$ is also model-derived from stomatal density, guard cell
length, assumed pore geometry, and conductance calculations. I think one paper
by the Sack and Buckley group explicitly describes anatomical $g_\mathrm{max}$
as a theoretical estimate of maximum stomatal diffusion capacity that cannot
usually be reached in practice, which is useful but not identical to observed
aperture. I would ask the authors to validate or reframe the variable.
Specifically, they should report posterior correlations between fitted \gi
and $\tau$ within individual curves, run a null simulation showing whether the
curve fitting procedure alone can induce a \gi–$\tau$ association, and
repeat the main analysis using alternative predictors such as absolute \gi,
observed maximum \gsw{}, \gmax{}, and \gi and \gmax{} as separate
predictors. If no direct pore aperture measurements exist, I would also
suggest using more cautious language such as "conductance fraction of
anatomical maximum" rather than "aperture."
:::

::: response
**Response:**

We thank the reviewer for raising this point because it revealed a mistake in our explanation of the methods. Figure 1 and our description incorrectly stated that the fitted interval began as soon as the RWR began. Our actual protocol was that we recorded \gsw{} before the VPD step and starting logging when \gsw{} returned to that point. We revised Figure 1a and the Materials and Methods (lines XXX-XXX). The reason for this protocol was that preliminary experiments showed the relationship between \gsw{} and $A$ analyzed in @muir_plasticity_2025 was "off" over the interval of \gsw{} between its maximum and when it returned to the pre-step value (the interval between the start of the RWR and when we started logging as depicted in Figure 1a). Our interpretation is that during the WWR \gsw{} increases faster than photochemistry can be upregulated, such that $A$ doesn't increase as much as you would expect for that amount of change in \gsw. This point is tangential to our analysis of stomatal kinetics, but we hope it clarifies the rationale. Hence, we only have a single of estimate of \fgmax{} where \gi{} is the pre-step \gsw{}, as depicted in the revised version of Figure 1a. We do not have data on maximum \gsw{} at the end of the WWR. In principle, if the relationship between guard cell turgor and \gsw{} (Figure 1d) within a leaf is fixed then we should be able to estimate the effect of \fgmax{} on $\tau$ either way, but we cannot evaluate this in more detail with the data we have. 

Note that figure S5 (previously figure S1) shows that estimated \gi is nearly perfectly correlated with measured \gsw{} at the initial point. Clarify that we started logging when we believed WWR was done, which was achieved by starting logging when \gsw reached the \gsw{} value before the VPD step change.

We investigated whether estimating \gi from the same nonlinear fit as $\tau$ and $\lambda$ could, by itself, introduce a spurious \gi-$\tau$ (and hence \fgmax{}-$\tau$) association with a null simulation. For every real curve, we simulated a synthetic \gsw{} trajectory using that curve's own measurement times and residual noise magnitude and its own fitted $\tau$ and $\lambda$, but with \gi and $g_f$ **fixed at their real median values for every curve** (so the true \fgmax{} has zero variance by construction and cannot be truly associated with $\tau$). We re-fit the same Weibull functional form to each simulated curve (via `nls()`, validated against the full Bayesian pipeline used for the real data on a 50-curve subsample: cor($\tau$) = `r fmt_cor(nls_vs_bayes$cor_tau)`, cor($\lambda$) = `r fmt_cor(nls_vs_bayes$cor_lambda)`) and computed the correlation between the *estimated* \fgmax{} and $\tau$. We repeated this 200 times to build a null distribution of this correlation. Because \gi and $g_f$ have no true variation in this simulation, any such correlation must be an artifact of the fitting procedure rather than biology.

The null-simulation correlations ranged from `r fmt_cor(min(null_sim_results$cor))` to `r fmt_cor(max(null_sim_results$cor))` (median `r fmt_cor(median(null_sim_results$cor))`) across the 200 replicates -- indistinguishable from zero -- whereas the real, observed correlation between \fgmax{} and $\tau$ is `r fmt_cor(real_cor_fgmax_tau$cor)` [`r fmt_cor(real_cor_fgmax_tau$cor_lower)`, `r fmt_cor(real_cor_fgmax_tau$cor_upper)`] (@fig-null-sim-fgmax-tau), far outside the null range (empirical $p$ = `r formatC(null_sim_p, format = "f", digits = 3)`). This indicates that the curve-fitting procedure alone cannot explain the observed \fgmax{}-$\tau$ association, and that it more plausibly reflects a real relationship in the data.

```{r}
#| label: fig-null-sim-fgmax-tau
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Null distribution of the correlation between estimated \\fgmax{} and estimated $\\tau$ across 200 simulated datasets in which the true \gi and $g_f$ are fixed at their real median values for every curve (so \\fgmax{} has no true relationship with $\\tau$ by construction), compared to the observed correlation in the real data (red line)."
knitr::include_graphics("../figures/null-sim-fgmax-tau.pdf")
```

While \gmax{} is theoretical to some extent, it does seem to be a good description of maximum \gsw{} in experimentally controlled *Arabidopsis* plants (Dow et al. 2014).

:::

---

## Comment R1.4 — Timing of \gi is ambiguous

::: revcomment
I also found the timing of \gi is ambiguous because the authors start
logging after the wrong way response. Figure 1A and lines 41–45 correctly
distinguish the WWR and RWR, but lines 248–249 state that data logging began
after the transient WWR. Then lines 273–278 say that fitted \gi is a good
proxy for conductance "at the time of the step change in humidity." These
statements do not seem consistent. If logging begins after the WWR, then \gi
is the initial conductance of the analyzed RWR segment and therefore not
necessarily conductance at the humidity step. Figure 1A actually illustrates
this distinction nicely with the VPD step occurring before the orange "log
data" interval. It would be great if author define \gi throughout as the
fitted initial conductance at the start of the analyzed RWR, unless pre-step
conductance was recorded and used. If pre-step steady state \gsw{} exists, the
authors could also consider computing a pre-step \fgmax{} and test whether it
predicts $\tau$, as that would be much more convincing mechanistically than
using a parameter estimated from the RWR itself.
:::

::: response
**Response:**

The reviewer is correct that Figure 1a and the accompanying description were internally inconsistent, and we thank the reviewer for catching this. This was an unintentional error in how we illustrated and described the timing of logging relative to the WWR; our data-collection protocol, curve fitting, and downstream analyses are unchanged.

To clarify: we deliberately began logging each curve once \gsw{} had declined back down to its pre-step value, rather than logging continuously from the moment of the humidity step or from the WWR peak. We used this design specifically so that the first logged value of \gsw{}, and hence the fitted \gi, would correspond to pre-step conductance rather than to some arbitrary point during the WWR overshoot. Consequently, \gi can accurately be described both ways the reviewer suggests: it is the fitted initial conductance at the start of the analyzed RWR segment, and by design it is also our estimate of pre-step conductance. Our \fgmax{} estimates ($g_i / g_\mathrm{max}$) already reflect pre-step conductance for this reason, so a separately computed pre-step \fgmax{} would not differ meaningfully from what we already report.

We revised Figure 1a and the Materials and Methods to describe this precisely, and moved the "Log data" marker in the figure to the point where \gsw{} returns to \gi, distinct from the start of the RWR itself, which begins where \gsw{} first starts to decline from its WWR peak. We also added a caveat to the Discussion noting this limitation.

One consequence of this design is that we did not log \gsw{} at its WWR maximum, so we lack a measurement of WWR magnitude and cannot directly test how it relates to $\tau$ or $\lambda$. By the logic of our approach, we would expect a larger-magnitude WWR to be associated with a slower, not faster, subsequent response. We agree that directly testing this relationship would be a valuable direction for future work, but it is not possible with the present dataset.
:::

---

## Comment R1.5 — High vs.\ low light confounded with measurement order

::: revcomment
Another confusing point is that the current high versus low measurement light
comparison is confounded with measurement order and prior stress history.
Lines 224–226 say that high light curves were always measured first, and lines
250–251 say the leaf was then acclimated to low light and high RH before
inducing the low light closure curve. This means the largest treatment effect
in the paper is also a fixed sequence effect with high light always first, and
low light always following a prior high light, low humidity closure event on
the same leaf. That matters because high measurement light strongly increases
both \fgmax{} and $\tau$ and the mediation analysis then interprets part of
the light effect as operating through \fgmax{}. I do not see how the current
design separates measurement light from order, recovery state, prior VPD
exposure, elapsed time, or time of day. It might be better to call this a
"measurement light/order condition" unless they can show that sequence effects
are negligible.
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R1.6 — "Causal mediation" language too strong

::: revcomment
Lines 339–347 describe "causal mediation analysis," and lines 445–455 report
mediated effects and Figure 7 presents path diagrams implying treatment effects
flowing through \fgmax{} to $\tau$. I understand why the authors did this but
the causal assumptions are too strong for the current design. \fgmax{} is not
randomized, it is estimated from the same curve as $\tau$, it is affected by
actual VPD trajectory and measurement order and it is likely related to
unmeasured hydraulic state. In the standard mediation literature, causal
mediation requires strong assumptions about mediator–outcome confounding usually
framed as sequential ignorability and sensitivity analysis is a central part of
the approach. It could be great if authors replace "causal mediation analysis"
with "statistical decomposition" or "path analysis" unless they add a DAG and
sensitivity analysis. The text could still say that treatment-associated changes
in $\tau$ are statistically consistent with an indirect path through \fgmax{},
but I would avoid saying that \fgmax{} "mediates plasticity" as a causal
conclusion.
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R1.7 — Model selection is more subjective than admitted

::: revcomment
I also found the model selection is more subjective than the present manuscript
currently admits. Lines 332–337 say that models were compared using LOOIC and
models within two $\Delta$LOOIC standard errors were considered plausible.
Table 2 shows six plausible models and the selected model is not the minimum
LOOIC model. Lines 430–435 then acknowledge that judgment calls were used,
which I appreciate. However, the code is more explicit than the manuscript:
`r/11_compare-models.R`, lines 74–84, says the top four model order changed
across reruns and that model 6 was chosen because it was "clearest to
interpret." That may be defensible but it should be stated plainly in the
manuscript. It would be great if authors can present focal estimates across all
plausible models not only the selected model, especially for \gcl{} → $\tau$,
\fgmax{} → $\tau$, \gcl{} → $\lambda$, and \fgmax{} → $\lambda$. They could
also consider reporting Pareto $\hat{k}$ diagnostics and Monte Carlo
uncertainty for PSIS-LOO, because PSIS-LOO model comparison relies on those
diagnostics. If you read the `loo` documentation where it emphasizes
Pareto-$k$, PSIS effective sample size and MCSE as reliability checks.
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R1.8 — Single-chain Bayesian diagnostics for curve-level fits

::: revcomment
The curve-level Bayesian diagnostics are not yet convincing enough for a
dataset built from 2,124 fitted nonlinear curves. Lines 285–289 state that
each curve was fit using 1,000 post-warmup draws from a single chain, with
Rhat $< 1.05$, bulk ESS $> 400$ and fewer than 10 divergent transitions. The
code confirms the single-chain curve fitting in `r/functions.R`, lines 8–18.
I am not fully convinced with this as the only diagnostic for thousands of
nonlinear curve fits especially because \gi, $g_f$, $\tau$, and $\lambda$
are likely correlated. Stan's diagnostics treat divergent transitions as a
warning that HMC may not be fully exploring the posterior and the usual
workflow relies on multiple chains to assess convergence. Perhaps no need for
the authors to refit all 2,124 curves with four chains if that is
computationally prohibitive, but it would be great for a stratified four-chain
validation subset including difficult curves, high-$\tau$ curves, high-\fgmax{}
curves, and low-$R^2$ curves. To show or supplement that the single-chain
pipeline gives essentially the same $\tau$, $\lambda$, and \gi estimates as
a more standard multi-chain fit.
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R1.9 — Undocumented guard cell length outlier exclusions

::: revcomment
I found in `r/01_join-data.R`, three guard cell length outlier individuals are
removed with a comment referring to `r/scratch - gcl outliers.R`. I do not see
this exclusion described in the manuscript and the referenced scratch file is
not included in the package. Because guard cell length is one of the two
central anatomical predictors, it would be great to see this reported. The
same script also uses a first-available fallback for stomatal anatomical data
when the LICOR and stomatal leaflet do not uniquely match
(`r/01_join-data.R`). That may be reasonable but it could be better to
document because it might add measurement error to \gmax{}, guard cell length
and therefore \fgmax{}.
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R1.10 — Figure 1D color labeling appears reversed

::: revcomment
In Figure 1D, the orange point is the low \fgmax{} point on the steep part of
the curve and the blue point is the high \fgmax{} point on the shallow part of
the curve? The figure labels themselves appear correct. However, the Figure 1
caption says the larger change occurs at low \fgmax{} "blue; steep tangent
slope" and high \fgmax{} is "orange; shallow tangent slope." I found that is
reversed relative to the plotted colors?
:::

::: response
**Response:**

The reviewer is correct. The in-figure labels and plotted colors in Figure 1d were always consistent (low \fgmax{} = orange/steep slope, high \fgmax{} = blue/shallow slope), but we had the colors swapped in the Figure 1 caption text. We corrected the caption to match the figure.
:::

---

# Reviewer 2

## Comment R2.1 — Abstract sentence unclear

::: revcomment
I am not sure this sentence reads clearly: "Leaves with small guard cells and
a lower \gop{} to \gmax{} ratio (\fgmax{}) closed faster, but explained
variation in kinetic parameters at different levels of biological
organization." Perhaps it is missing 'both'? "Leaves with small guard cells
and a lower \gop{} to \gmax{} ratio (\fgmax{}) closed faster, but both
explained variation in kinetic parameters at different levels of biological
organization."
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R2.2 — Mechanistic follow-up on turgor-responsive window

::: revcomment
This is really interesting and links back to early work on turgor maintenance
and speed: "Smaller stomata can be speedier, but only if stomata are held at
an aperture where they are responsive to changing turgor pressure." How could
the authors mechanistically follow this up? Using anatomical measurements of
$g_\mathrm{smax}$ as a benchmark for comparing to operational \gsw{} has some
drawbacks — e.g.\ over-estimation. I think some of this is covered in the
discussion.
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R2.3 — Wrong-way response magnitude and \fgmax{}

::: revcomment
The authors have chosen species with 'wrong way response' to VPD — do they
take into account or explore the magnitude of the initial rise in \gsw{} and
how this might relate to lag time or the \fgmax{} relationship (the magnitude
of the 'hump' in Figure 1a)?
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R2.4 — Compensatory aperture in adaxial-blocked leaves (L403)

::: revcomment
L403 — stomatal aperture was more open in leaves where gas exchange through
adaxial stomata was blocked — is this interesting, do the authors think it is
compensatory?
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R2.5 — Figure 1B legend format

::: revcomment
Fig 1b — Personal preference; although I appreciate the legend is correct, it
might be clearer to make the large and small time constants a rectangle of the
appropriate colour or dot of the colour? When I look at the figure I focus in
on the format of the line rather than the colour which distracts from
understanding the figure quickly.
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::
