---
title: "Response to Reviewers"
subtitle: |
  NPH-MS-2026-57029\
  *Guard cell size and pore aperture influence stomatal closure kinetics*\
  Muir & Lim
format:
  pdf:
    toc: false
    number-sections: false
    indent: false
    fontsize: 11pt
    geometry: margin=1in
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
header-includes: |
  \usepackage{xcolor}
  \usepackage{mdframed}
  \usepackage{setspace}
  \usepackage{placeins}
  \usepackage{siunitx}
  \renewcommand{\thefigure}{R\arabic{figure}}
  \definecolor{responseblue}{RGB}{70, 130, 180}
  \definecolor{commentgray}{RGB}{90, 90, 90}
  \newenvironment{response}
    {\color{responseblue}\setlength{\parskip}{0.5em}}
    {}
  \newenvironment{revcomment}
    {\color{commentgray}\itshape\begin{mdframed}[linecolor=commentgray,linewidth=0.5pt,innerleftmargin=8pt,innerrightmargin=8pt,innertopmargin=6pt,innerbottommargin=6pt]}
    {\end{mdframed}}
  \newcommand{\fgmax}{$f_\mathrm{gmax}$}
  \newcommand{\gcl}{$l_\mathrm{gc}$}
  \newcommand{\gf}{$g_\mathrm{f}$}
  \newcommand{\gi}{$g_\mathrm{i}$}
  \newcommand{\gmax}{$g_\mathrm{max}$}
  \newcommand{\gop}{$g_\mathrm{op}$}
  \newcommand{\gsw}{$g_\mathrm{sw}$}
bibliography: solanum-kinetics.bib
csl: https://www.zotero.org/styles/new-phytologist
---

We thank the editor and both reviewers for their thoughtful and constructive
comments. We address each point below. Reviewer comments appear in gray boxes;
our responses appear in blue.

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(cowplot)
library(dplyr)
library(ggplot2)
library(glue)
library(kableExtra)
library(magrittr)
library(readr)
library(stringr)

source("../r/functions.R")

df_comparison = read_rds("../objects/tbl-comparison.rds")
comp_pars_vpd = read_rds("../objects/tbl-comparison-vpd.rds")
final_vpd_effect = read_rds("../objects/tbl-final-vpd-effect.rds")
cor_tau_rate = read_rds("../objects/cor-tau-vpd-rate.rds")

fmt_est = function(.df, .parameter, .model) {
  row = .df[grepl(.parameter, .df$parameter, fixed = TRUE) & .df$model == .model, ]
  glue(
    "{formatC(row$estimate, format = 'f', digits = 2)} ",
    "[{formatC(row$lower, format = 'f', digits = 2)}, ",
    "{formatC(row$upper, format = 'f', digits = 2)}]"
  )
}

gi_orig = fmt_est(comp_pars_vpd, "italic(g)[i]", "Original")
gi_vpd = fmt_est(comp_pars_vpd, "italic(g)[i]", "VPD-adjusted")
gcl_orig = fmt_est(comp_pars_vpd, "phylogenetic~corr", "Original")
gcl_vpd = fmt_est(comp_pars_vpd, "phylogenetic~corr", "VPD-adjusted")

null_sim_results = read_rds("../objects/null-sim-fgmax-tau.rds")
real_cor_fgmax_tau = read_rds("../objects/real-cor-fgmax-tau.rds")
null_sim_p = read_rds("../objects/null-sim-empirical-p.rds") |>
  equals(0) |>
  ifelse(glue("empirical $P < {x}$", x = formatC(1 / nrow(null_sim_results), format = "f", digits = 3)
), glue("empirical $P = {x}$", x = formatC(null_sim_p, format = "f", digits = 3)))
nls_vs_bayes = read_rds("../objects/nls_vs_bayes.rds")

ct_tau = cor.test(nls_vs_bayes$logtau_brms, nls_vs_bayes$logtau_nls)
ct_lambda = cor.test(nls_vs_bayes$loglambda_brms, nls_vs_bayes$loglambda_nls)

fmt_cor = function(x) {
  if (abs(x) < 0.005) {
    # Convert to scientific notation in LaTeX
    sci_format <- format(x, scientific = TRUE, digits = 2)
    # Parse out the mantissa and exponent
    parts <- strsplit(sci_format, "e")[[1]]
    mantissa <- as.numeric(parts[1])
    exponent <- as.numeric(parts[2])
    glue("$", mantissa, " \\times 10^{{", exponent, "}}$")
  } else {
    formatC(x, format = "f", digits = 2)
  }
}

final_vpd_on_tau = glue(
  "{formatC(final_vpd_effect$estimate, format = 'f', digits = 2)} ",
  "[{formatC(final_vpd_effect$lower, format = 'f', digits = 2)}, ",
  "{formatC(final_vpd_effect$upper, format = 'f', digits = 2)}]"
)

mediation_sens = read_rds("../objects/mediation-sensitivity-rhostar.rds")

fmt_rho = function(param) {
  row = mediation_sens$rho_summary[mediation_sens$rho_summary$parameter == param, ]
  glue(
    "{formatC(row$estimate, format = 'f', digits = 2)} ",
    "[{formatC(row$lower, format = 'f', digits = 2)}, ",
    "{formatC(row$upper, format = 'f', digits = 2)}]"
  )
}

rho_hat_fmt = fmt_rho("rho_hat (estimated)")
rho_star_fmt = fmt_rho("rho_star (breakdown)")
prob_below_breakdown_fmt = formatC(100 * mediation_sens$prob_below_breakdown, format = "f", digits = 0)

gcl_switch_summary = read_rds("../objects/gcl_switch_summary.rds")

curve_vpd_summary = read_rds("../objects/curve-vpd-summary.rds")
cor_median_final_vpd = cor(curve_vpd_summary$median_vpd, curve_vpd_summary$final_vpd,
                            use = "complete.obs")

tbl_loo_diagnostics = read_rds("../objects/tbl-loo-diagnostics.rds")

```

---

# Reviewer 1

## Comment R1.1 — Realized VPD as a covariate

::: revcomment
It seems like the realized VPD stimulus reported varies systematically among
treatments and this might needs to be treated as part of the analysis rather
than as a supplementary caveat, as in Lines 238–249 are unusually candid and
useful, as the authors state that scrubbing incoming water vapor "did not
standardize chamber air humidity because leaf transpiration introduced water
vapor," and that RH was systematically higher at high measurement light, in
sun grown plants and in amphistomatous measurements because \gsw{} was greater
in those conditions. Table S3 then shows that the mean VPD during measurements
ranged from 2.50 kPa in sun/high/amphi curves to 2.92 kPa in
shade/low/pseudohypo curves. I think this matters because the actual forcing
variable in the experiment is not simply the experimental treatment label, but
also is the VPD trajectory experienced by the leaf. If leaves with higher
conductance both have higher \fgmax{} and also humidify the chamber more
strongly, then \fgmax{} is partly entangled with the realized stimulus. I would
love to see the authors include curve level VPD covariates in the model such as
median VPD during the fitted interval, final VPD if available, VPD step
magnitude, and perhaps the rate at which VPD changed. At minimum, should show
whether the \fgmax{} → $\tau$ effect remains after accounting for actual VPD
exposure, and ideally they should run a matched VPD sensitivity analysis
excluding curves outside a narrow VPD range. I would be careful describing the
VPD treatment as though the desiccant step created a common perturbation across
treatments unless this is demonstrated. This distinction is important because
VPD is a primary driver of stomatal conductance and plant water relations, and
the physiologically relevant stimulus is the leaf to air evaporative gradient
not simply the inlet humidity setting.
:::

::: response

<!-- TODO:
- put in correct Figure numbers for main text
-->
 
**Response:**

We added Notes S1 to the Supporting Information to address the potential for variable VPD stimuli to confound our results, and updated the Materials and Methods to point to it. Note that in response Comment R1.7, we found that \gi{} is more strongly associated with $\tau$ than \fgmax. Hence, we discuss whether effects \gi{} and \gcl{} on $\tau$ are robust once variation in VPD is incorporated into the analysis. We find that both effects are robust but slightly weaker once final VPD is included as a covariate in the selected model. The fixed effect of \gi{} on $\log \left( \tau \right)$ (original: `r gi_orig`; VPD-adjusted: `r gi_vpd`) and the phylogenetic correlation between \gcl{} and $\tau$ (original: `r gcl_orig`; VPD-adjusted: `r gcl_vpd`) were both slightly weaker after adding final VPD as a covariate of $\tau$ and $\lambda$ (@fig-vpd-comparison; Figure SXXX in the Supporting Information). One caveat is that we do not have data on the initial rate of change in VPD, which limits our ability to evaluate whether this influenced stomatal kinetic responses; we added a recommendation for future research on this point at the end of Notes S1.

We extracted the leaf-to-air VPD trajectory for every curve from the logged humidity data and summarized it with two curve-level covariates: the VPD at the end of the fitted interval ("final VPD") and the rate at which VPD approached that final value ("VPD saturation rate"). We considered adding
median VPD and VPD step magnitude as well, as the reviewer suggested. However, median VPD and final VPD are highly correlated ($r = `r formatC(cor_median_final_vpd, format = "f", digits = 2)`$; @fig-vpd-pairs a; Figure SXXX in the Supporting Information), so we chose to use the final VPD. Final VPD contains the same information as the VPD step magnitude since all leaves were started at the same RH and Tleaf, hence starting VPD was approximately constant.

We attempted to incorporate the rate of VPD change as a covariate, but determined this was not useful. Unfortunately, we did not log data during the WWR when the initial step change in VPD occurred. We instead fit a saturating exponential, $\mathrm{VPD}(t) = B + (R_0 - B)\exp(-kt)$, to each curve's VPD trajectory (median $R^2 = 0.99$) and use the rate constant $k$ ("VPD saturation rate"). This rate constant is only weakly correlated with final VPD ($r = 0.31$; @fig-vpd-pairs c). However, $k$ is strongly correlated with $\tau$ ($r = `r formatC(cor_tau_rate, format = "f", digits = 2)`$ on a log-log scale; @fig-tau-vpd-rate) because it is mostly caused by decreasing $g_\mathrm{sw}$. We can see this clearly by comparing the rate of change in the water vapor concentration of the reference airstream, upstream of the leaf ($W_\mathrm{r}$), versus that of the sample airstream, which reflects humidity inside the chamber ($W_\mathrm{s}$). If the change in VPD is caused by $g_\mathrm{sw}$, we expect a positive correlation between the rate of VPD change and the rate of $W_\mathrm{s}$ change, which is exactly what we see (@fig-vpd-h2o-rate a; Figure SXXX in the Supporting Information). In contrast, the rate of $W_\mathrm{r}$ change is uncorrelated with the rate of VPD change (@fig-vpd-h2o-rate b; Figure SXXX in the Supporting Information). Hence, it would not be statistically valid to include the rate of VPD change as a covariate in predicting $\tau$ since both are caused by the same underlying variable, the rate of change in $g_\mathrm{sw}$.

To directly test whether the realized VPD stimulus confounds the \gi{} and \gcl{} effects on $\tau$, we refit our selected model with final VPD added as a covariate of $\tau$ and $\lambda$. As reported above, the \gi{} and \gcl{} effects were slightly weaker when final VPD was included as a covariate (@fig-vpd-comparison). Final VPD itself had a negative fixed effect on $\log \tau$ (`r final_vpd_on_tau`), consistent with higher realized VPD accelerating stomatal closure.

```{r}
#| label: fig-vpd-comparison
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The \\gi{} effect on $\\tau$ and the phylogenetic correlation between \\gcl{} and $\\tau$ are robust to accounting for final VPD. Posterior means and 95\\% credible intervals for the fixed effect of $\\log \\left( g_\\mathrm{i} \\right)$ on $\\log \\left( \\tau \\right)$ (left) and the phylogenetic correlation between $\\log \\left( l_\\mathrm{gc} \\right)$ and $\\log \\left( \\tau \\right)$ (right), comparing the original model (grey) to a model refit with final VPD added as a covariate (pink)."
knitr::include_graphics("../figures/vpd-comparison.pdf")
```

```{r}
#| label: fig-vpd-pairs
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Pairwise relationships among curve-level VPD covariates: (a) median vs. final VPD, (b) median VPD vs. VPD saturation rate, (c) final VPD vs. VPD saturation rate. Each point is one humidity response curve, colored by leaf type; lines are ordinary least-squares fits within leaf type. VPD saturation rate is the rate constant of a saturating exponential fit to the VPD trajectory during the fitted interval."
knitr::include_graphics("../figures/vpd-pairs.pdf")
```

```{r}
#| label: fig-tau-vpd-rate
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The stomatal closure time-constant $\\tau$ is positively correlated with the inverse of the VPD saturation rate, $k^{-1}$ (both log scale; $r = `r formatC(cor_tau_rate, format = 'f', digits = 2)`$), consistent with $k$ being largely driven by the same $g_\\mathrm{sw}$-decline signal that determines $\\tau$, rather than an independent property of the VPD stimulus. Each point is one humidity response curve."
knitr::include_graphics("../figures/tau-vpd-rate.pdf")
```

```{r}
#| label: fig-vpd-h2o-rate
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "The VPD saturation rate is driven by the sample airstream, not the reference airstream. (a) VPD saturation rate vs. the saturation rate of $W_\\mathrm{s}$ (sample cell water vapor concentration, which has equilibrated with the leaf); the dashed line is the 1:1 line. (b) VPD saturation rate vs. the saturation rate of $W_\\mathrm{r}$ (reference cell water vapor concentration, upstream of the leaf). Each point is one humidity response curve; rate constants come from the same saturating-exponential model used for VPD."
knitr::include_graphics("../figures/vpd-h2o-rate.pdf")
```

:::

```{=latex}
\FloatBarrier
```

---

## Comment R1.2 — Mislabeled VPD summary statistic

::: revcomment
Related to the previous point, the manuscript appears to call the Table S3
values "final RH" and "final VPD," but I found the code and caption indicate
something different. At lines 245 to 247, the text says that "the final RH"
and "VPD" values in Table S3 elicited rapid stomatal response. However, the
Table S3 caption says the values are calculated as the median RH and VPD
across time points within each curve and then averaged among curves. The code
confirms this: `r/09_make-tbl-vpd.R`, lines 7–14, calculates median RH and
VPD per curve, and lines 32–40 average those medians across treatment
combinations. That is not the same as final RH/VPD nor necessarily the VPD
step magnitude. It would be great if author can correct the wording and, more
importantly, to report the full VPD trajectory more transparently, with
pre-step VPD, VPD at the start of logging, median VPD during the fitted RWR,
final VPD, and ideally the change in VPD over time, because the paper
interprets variation in $\tau$ as intrinsic closure kinetics, while the
stimulus itself may vary across the same treatments used to explain $\tau$.
:::

::: response

<!-- DONE -->

**Response:**

We thank the reviewer for catching the discrepancy between what we reported in Table S3 versus the actual content. At noted in the response to the previous comment, the median and final RH and VPD are highly correlated, so this mistake did not affect the interpretation. We revised Table S3 to report the average RH and VPD at three time points: when logging started, when logging finished, and the median during the logging interval. We also added the curve-level values for these variables to Table S1. We did not record the pre-step VPD but it was nearly identical for every leaf because we set leaf temperature to $\SI{25}{\degreeCelsius}$ and RH to 70% to standardize starting conditions. This is equivalent to $\mathrm{VPD} \approx \SI{0.82}{\kPa}$, as mentioned in the Materials and Methods. As mentioned in the previous response, we also do not have data on the initial rate of VPD increase.

:::

---

## Comment R1.3 — \fgmax{} is not an independent aperture measurement

::: revcomment
I like \fgmax{} as a quite intriguing predictor but it is not an independent
direct measure of stomatal aperture. The manuscript motivates \fgmax{} as
proportional to aperture relative to maximum aperture at lines 115–133,
defines it as $g_i / g_\mathrm{max}$ at lines 273–279 and then interprets
higher \fgmax{} as stomata being more open at lines 397–404 and as a
mechanistic driver of slower closure at lines 506–513. I think the hypothesis
is interesting, but the current evidence is not enough to treat \fgmax{} as a
direct aperture measurement. The numerator, \gi, is estimated from the same
nonlinear curve that also gives $\tau$ and $\lambda$, so parameter covariance
in the Weibull fit could generate an association between \gi{} and $\tau$ even
if the underlying biological relationship is weaker. The denominator
$g_\mathrm{max}$ is also model-derived from stomatal density, guard cell
length, assumed pore geometry, and conductance calculations. I think one paper
by the Sack and Buckley group explicitly describes anatomical $g_\mathrm{max}$
as a theoretical estimate of maximum stomatal diffusion capacity that cannot
usually be reached in practice, which is useful but not identical to observed
aperture. I would ask the authors to validate or reframe the variable.
Specifically, they should report posterior correlations between fitted \gi{}
and $\tau$ within individual curves, run a null simulation showing whether the
curve fitting procedure alone can induce a \gi–$\tau$ association, and
repeat the main analysis using alternative predictors such as absolute \gi,
observed maximum \gsw{}, \gmax{}, and \gi{} and \gmax{} as separate
predictors. If no direct pore aperture measurements exist, I would also
suggest using more cautious language such as "conductance fraction of
anatomical maximum" rather than "aperture."
:::

::: response
**Response:**

<!-- TODO:
- check final numbers once vpd model is run a final time
- put in correct Figure numbers for main text
- Make sure that text and figures here align with main text
- Address issue about "aperture" language throughout
- Are there any particular places to point to where the new analysis has altered interpretation?
 -->

We thank the reviewer for raising this point because it revealed a mistake in our explanation of the methods that is also relevant to Comment R1.4 below. Figure 1a and parts of the written description incorrectly stated that the fitted interval began as soon as the RWR began. To clarify, we deliberately began logging each curve once \gsw{} had declined back down to its pre-step value. We did this by recording the pre-step \gsw{} and starting the log protocol when \gsw{} returned to value after the WWR. The reason for this protocol was that preliminary experiments showed the relationship between \gsw{} and $A$ analyzed in @muir_plasticity_2025 was "off" over the interval of \gsw{} between its maximum and when it returned to the pre-step value (the interval between the start of the RWR and when we started logging as depicted in Figure 1a). Our interpretation is that during the WWR \gsw{} increases faster than photochemistry can be upregulated, such that $A$ doesn't increase as much as you would expect for that amount of change in \gsw. This point is tangential to our analysis of stomatal kinetics, but we hope it clarifies the rationale. Hence, we only have a single of estimate of \fgmax{} where \gi{} is the pre-step \gsw{}, as depicted in the revised version of Figure 1a. We do not have data on maximum \gsw{} at the end of the WWR. \gi{} is therefore also the maximum observed \gsw{} because we allowed leaves to acclimate under conditions (high light, low VPD) conducive to high \gsw. In principle, if the relationship between guard cell turgor and \gsw{} (Figure 1d) within a leaf is fixed then we should be able to estimate the effect of \fgmax{} on $\tau$ either regardless of where along the response curve we start, but we cannot evaluate this in more detail with the data we have. 

We revised Figure 1a and the Materials and Methods (lines XXX-XXX) to describe this precisely, and moved the "Log data" marker in the figure to the point where \gsw{} returns to \gi, distinct from the start of the RWR itself, which begins where \gsw{} first starts to decline from its WWR peak. We also added a caveat to the Discussion noting this limitation.

We also added simulations demonstrating that the association between \gi, \fgmax{}, and $\tau$ is unlikely to be an artifact of simultaneously estimating \gi{} and $\tau$. For every real curve, we simulated a synthetic \gsw{} trajectory using that curve's own measurement times, residual variance, and fitted $\tau$ and $\lambda$ values, but with fitted values of \gi{} and \gf{} randomly swapped with another curve. Hence, synthetic curves retain the observed associations between $\tau$ and $\lambda$ from curve $i$ and the observed associations between \gi{} and \gf{} from curve $j$. Since $i$ and $j$ are chosen at random, there is no actual association between \gi{} and kinetic parameters among the synthetic curves. Since the denominator in \fgmax{} is estimated independently from stomatal anatomy, it is sufficient to evaluate a spurious association between \gi{} and $\tau$. For each synthetic curve, we re-fit the same Weibull functional form to each simulated curve (via `nls()`, validated against the full Bayesian pipeline used for the real data: cor($\tau$) = `r fmt_cor(ct_tau$estimate)`, cor($\lambda$) = `r fmt_cor(ct_lambda$estimate)`) and computed the correlation between the *estimated* \gi{} and $\log \left( \tau \right)$. We repeated this `r prettyNum(nrow(null_sim_results), big.mark = ",")` times to build a null distribution of this correlation. 

The null-simulation correlations ranged from `r fmt_cor(min(null_sim_results$cor))` to `r fmt_cor(max(null_sim_results$cor))` (median `r fmt_cor(median(null_sim_results$cor))`) across the `r prettyNum(nrow(null_sim_results), big.mark = ",")` synthetic curves -- indistinguishable from zero -- whereas the real, observed correlation between $\log \left( g_\mathrm{i} \right)$ and $\log \left( \tau \right)$ is `r fmt_cor(real_cor_fgmax_tau$cor)` [`r fmt_cor(real_cor_fgmax_tau$cor_lower)`, `r fmt_cor(real_cor_fgmax_tau$cor_upper)`] (@fig-null-sim-fgmax-tau; Figure SXXX in the Supporting Information), far outside the null range (`r null_sim_p`). This indicates that the curve-fitting procedure alone cannot explain the observed \gi-$\tau$ association, and that it more plausibly reflects a real relationship in the data.

```{r}
#| label: fig-null-sim-fgmax-tau
#| echo: false
#| message: false
#| warning: false
#| fig-cap: !expr paste("Null distribution of the correlation between estimated \\gi{} and estimated $\\tau$ across", prettyNum(nrow(null_sim_results), big.mark = ","), "simulated datasets in which the true \\gi{} and \\gf{} are fixed at their real median values for every curve (so \\gi{} has no true relationship with $\\tau$ by construction), compared to the observed correlation in the real data (red line).")
#| 
knitr::include_graphics("../figures/null-sim-fgmax-tau.pdf")
```

While \gmax{} is theoretical to some extent, it does seem to be a good description of maximum \gsw{} in experimentally controlled *Arabidopsis* plants [@dow_integrated_2014]. We added this in the Materials and Methods on lines XXX.

To address the last part of the reviewer's comment, we substantially revised the model set we considered and this led to significant changes in our conclusions. The reviewer is correct that we do not have direct estimates of stomatal aperture and we revised the manuscript to reflect that our inferences are indirect. We also tested for independent effects of \gi{} and \gmax{} that are described in response to Comment R1.7 below. The new approach revealed that \gi{} is more closely associated with kinetic parameters than \gmax. Therefore, both greater density and/or aperture may slow the rate of closure. We have revised the Results and Discussion in several places to reflect this change in our interpretation.
:::

---

## Comment R1.4 — Timing of \gi{} is ambiguous

::: revcomment
I also found the timing of \gi{} is ambiguous because the authors start
logging after the wrong way response. Figure 1A and lines 41–45 correctly
distinguish the WWR and RWR, but lines 248–249 state that data logging began
after the transient WWR. Then lines 273–278 say that fitted \gi{} is a good
proxy for conductance "at the time of the step change in humidity." These
statements do not seem consistent. If logging begins after the WWR, then \gi{}
is the initial conductance of the analyzed RWR segment and therefore not
necessarily conductance at the humidity step. Figure 1A actually illustrates
this distinction nicely with the VPD step occurring before the orange "log
data" interval. It would be great if author define \gi{} throughout as the
fitted initial conductance at the start of the analyzed RWR, unless pre-step
conductance was recorded and used. If pre-step steady state \gsw{} exists, the
authors could also consider computing a pre-step \fgmax{} and test whether it
predicts $\tau$, as that would be much more convincing mechanistically than
using a parameter estimated from the RWR itself.
:::

::: response

<!-- DONE -->

**Response:**

Please see response to Comment R1.3 above.

:::

---

## Comment R1.5 — High vs. low light confounded with measurement order

::: revcomment
Another confusing point is that the current high versus low measurement light
comparison is confounded with measurement order and prior stress history.
Lines 224–226 say that high light curves were always measured first, and lines
250–251 say the leaf was then acclimated to low light and high RH before
inducing the low light closure curve. This means the largest treatment effect
in the paper is also a fixed sequence effect with high light always first, and
low light always following a prior high light, low humidity closure event on
the same leaf. That matters because high measurement light strongly increases
both \fgmax{} and $\tau$ and the mediation analysis then interprets part of
the light effect as operating through \fgmax{}. I do not see how the current
design separates measurement light from order, recovery state, prior VPD
exposure, elapsed time, or time of day. It might be better to call this a
"measurement light/order condition" unless they can show that sequence effects
are negligible.
:::

::: response

<!-- TODO:
- put in correct lines numbers of changes
-->

**Response:**

We choose this design because the primary goals of the study were to compare amphi and pseudohypo curves on the same leaf within the same treatment to estimate "amphistomy advantange" [@muir_plasticity_2025]. Preliminary studies indicated the order did not affect amphistomy advantange but that it was much faster to go from the high light to low light compared to the reverse. Therefore, we opted to increase our throughput and sample size at the potential cost of some confounding between measurement order and treatment effect. Unfortunately, we did not observe whether the measurement order affected stomatal kinetics. Hence, we cannot say for sure whether order, recovery state, or prior VPD exposure confounds measurement light intensity. We controlled for time of day by running curves throughout the day and randomizing the order among accessions and growth light treatments. However, we think the bias induced by measurement order is likely small compared to the very large effect of measurement light intensity on \gi{} (+77% for shade plants; +139% for sun plants):

| Growth light | Measurement: low | Measurement: high |
| --- | --- | --- |
| shade | 0.209 | 0.370 |
| sun | 0.265 | 0.634 |

Table: Average \gi{} in different treatment combinations across the amphi leaf type curves.

We have already mentioned that this analysis is opportunistic and therefore some of the experimental design choices are sub-optimal for the purpose of study stomatal kinetics. We addressed the reviewer's concerns about measurement order more specifically by:

1. Mentioning the potential for measurement order to confound effects of measurement light treatments (Materials and Methods lines XXX-XXX).

> "As a consequence, measurement light intensity is confounded with measurement order in our design: low light-intensity curves were always measured after high light-intensity curves on the same leaf, following a preceding low-humidity closure event, so we cannot fully separate an effect of light intensity per se from order, recovery state, or prior VPD exposure (see Discussion)."

2. Advising caution in interpreting effects of measurement light treatment and mediation analysis (Discussion lines XXX-XXX).

> "This comparison should be interpreted cautiously, however, because measurement light intensity was confounded with measurement order: high light curves always preceded low light curves on the same leaf, following a prior low-humidity closure event (see Materials and Methods). We cannot rule out that some of the apparent effect of measurement light intensity on \fgmax{} and $\tau$, and the corresponding indirect path through \fgmax{} in our mediation analysis ([@fig-mediation]), reflects order, incomplete recovery, or prior VPD exposure rather than measurement light intensity itself. This concern applies specifically to the measurement light comparison and not to growth light intensity or leaf type, which were randomized or alternated, respectively."

3. Advising that future studies randomize or alternate measurement order to avoid this issue (Discussion lines XXX).

> "Future studies should also randomize or alternate the order in which repeated within-leaf treatments (such as measurement light intensity) are applied, so that treatment effects can be separated from measurement order, recovery state, and prior stress history, which were confounded in our design."

4. Adding that the average time between high and low measurement light intensity curves was 48.9 minutes (lines XXX).

> "...on average 48.9 minutes after the start of the high-light curve."

:::

---

## Comment R1.6 — "Causal mediation" language too strong

::: revcomment
Lines 339–347 describe "causal mediation analysis," and lines 445–455 report
mediated effects and Figure 7 presents path diagrams implying treatment effects
flowing through \fgmax{} to $\tau$. I understand why the authors did this but
the causal assumptions are too strong for the current design. \fgmax{} is not
randomized, it is estimated from the same curve as $\tau$, it is affected by
actual VPD trajectory and measurement order and it is likely related to
unmeasured hydraulic state. In the standard mediation literature, causal
mediation requires strong assumptions about mediator–outcome confounding usually
framed as sequential ignorability and sensitivity analysis is a central part of
the approach. It could be great if authors replace "causal mediation analysis"
with "statistical decomposition" or "path analysis" unless they add a DAG and
sensitivity analysis. The text could still say that treatment-associated changes
in $\tau$ are statistically consistent with an indirect path through \fgmax{},
but I would avoid saying that \fgmax{} "mediates plasticity" as a causal
conclusion.
:::

::: response

<!-- TODO:
- put in correct Figure numbers for main text
-->

**Response:**

We agree and have replaced "causal mediation analysis" with "path analysis" throughout the manuscript. Note that in response to Comment R1.3 and R1.7, we found stronger evidence for \gi{} rather than \fgmax{} affecting $\tau$. Therefore, we reran path analysis using \gi{} in place of \fgmax{} as described in the revised "Path analysis to test for plastic effects" section. We no longer say that \gi{} "mediates plasticity" as a causal conclusion. We now state only that treatment-associated changes in $\tau$ are statistically consistent with an indirect path through \gi, alongside a direct path, exactly as the reviewer suggested.

We also drew the DAG the reviewer requested requested (@fig-mediation-dag; Figure XXX in the manuscript), showing the assumed causal structure (treatment $\to$ \gi{} $\to$ $\tau$ and treatment $\to$ $\tau$ directly) alongside an unmeasured confounder $U$ representing exactly the kind of variable the reviewer raised (hydraulic status, measurement order, realized VPD trajectory) that would violate sequential ignorability if it affected both \gi{} and $\tau$.

```{r}
#| label: fig-mediation-dag
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Assumed causal structure of the \\gi{}-$\\tau$ path analysis. Treatment has a direct effect on $\\tau$ (bottom arrow) and an indirect effect acting through \\gi{} (upper path). An unmeasured confounder $U$ (dashed arrows) affecting both \\gi{} and $\\tau$ would violate the sequential ignorability assumption required to interpret the indirect path causally."
knitr::include_graphics("../figures/mediation-dag.pdf")
```

To directly address sequential ignorability, we ran a sensitivity analysis in Notes S3 following the general framework of @imai_general_2010, parameterized by $\rho = \mathrm{Cor}(e_{g_\mathrm{i}}, e_\tau)$, the residual correlation between the \gi{} and $\tau$ equations. Under a violation of sequential ignorability with residual correlation $\rho$, the bias-adjusted mediator-on-outcome coefficient is $\beta_2(\rho) = \hat\beta_2 - \rho \cdot \sigma_\tau / \sigma_{g_\mathrm{i}}$, so the average causal mediation effect (ACME) is $\mathrm{ACME}(\rho) = \gamma_1 \cdot \beta_2(\rho)$, and the "breakdown point" $\rho^*$ at which the ACME is exactly zero is $\rho^* = \hat\beta_2 \cdot \sigma_{g_\mathrm{i}} / \sigma_\tau$. In these equations, $\beta_2$ is the effect of \gi{} on $\tau$ and $\gamma_1$ is the effect of treatments on \gi.

Our selected model already estimates the residual correlation between the \gi{} and $\tau$ equations ($\rho_\mathrm{estimated}$, from `set_rescor(TRUE)`), because we fit all five responses (\gcl, \gi, \gmax, $\tau$, $\lambda$) jointly. We treat $\hat\beta_2$ as the "naive" coefficient in the sensitivity framework above, and the model's own estimated residual correlation as an estimate of $\rho$.

We found $\rho_\mathrm{estimated}$ = `r rho_hat_fmt` and $\rho^*$ = `r rho_star_fmt` (@fig-mediation-sensitivity, Figure XXX in the revised manuscript). These posteriors do not substantially overlap and the posterior probability that the estimated residual correlation is smaller in magnitude than the breakdown point is nearly `r prob_below_breakdown_fmt`%. In other words, the plastic mediation effects of \gi{} on $\tau$ are probably not explained by unmeasured confounders. Nevertheless, we took the reviewer's suggestion to avoid strong causal claims because the mediator variable \gi{} was not experimentally controlled and could plausibly be related to $\tau$ through confounding variables, though our analysis does not support this conclusion.

```{r}
#| label: fig-mediation-sensitivity
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Sensitivity of the average causal mediation effect (ACME) of each treatment on $\\log \\tau$, acting through \\fgmax{}, to an assumed residual correlation $\\rho$ between the \\fgmax{} and $\\tau$ equations. The vertical dotted line is the model's own estimated residual correlation ($\\rho_\\mathrm{estimated}$, shaded band = 95\\% credible interval); the vertical solid line is the breakdown point $\\rho^*$ at which the ACME is zero."
knitr::include_graphics("../figures/mediation-sensitivity.pdf")
```

:::

---

## Comment R1.7 — Model selection is more subjective than admitted

::: revcomment
I also found the model selection is more subjective than the present manuscript
currently admits. Lines 332–337 say that models were compared using LOOIC and
models within two $\Delta$LOOIC standard errors were considered plausible.
Table 2 shows six plausible models and the selected model is not the minimum
LOOIC model. Lines 430–435 then acknowledge that judgment calls were used,
which I appreciate. However, the code is more explicit than the manuscript:
`r/11_compare-models.R`, lines 74–84, says the top four model order changed
across reruns and that model 6 was chosen because it was "clearest to
interpret." That may be defensible but it should be stated plainly in the
manuscript. It would be great if authors can present focal estimates across all
plausible models not only the selected model, especially for \gcl{} → $\tau$,
\fgmax{} → $\tau$, \gcl{} → $\lambda$, and \fgmax{} → $\lambda$. They could
also consider reporting Pareto $\hat{k}$ diagnostics and Monte Carlo
uncertainty for PSIS-LOO, because PSIS-LOO model comparison relies on those
diagnostics. If you read the `loo` documentation where it emphasizes
Pareto-$k$, PSIS effective sample size and MCSE as reliability checks.
:::

::: response

<!-- TODO:
- put in correct Figure and Table and Line numbers for main text
- cite specific instances where Results and interpretation changed (overlap with TODO for R1.3)
-->

**Response:**

The phrase "clearest to interpret" was informal shorthand (CDM) for the fuller explanation in the text, so there is no discrepancy to correct. To make model selection less subjective, we now describe explicit criteria in the revised "Model selection and parameter estimation of individual-level effects" section. We first narrowed the candidate set using LOOIC, then evaluated the remaining models on the focal parameters addressing our hypotheses — individual-level and phylogenetic effects of anatomical traits on kinetic traits. Among these, we selected the model with the highest proportion of significant focal parameters (95% CIs excluding 0), regardless of whether the direction matched our hypotheses. This criterion favors more parsimonious models, particularly ones without additional non-significant terms, and makes the selected model easier to interpret. This criterion is somewhat arbitrary and not based on formal statistical inference. We therefore also discuss alternative interpretations supported by other plausible models. In the revised model comparison, this approach happened to select the model with the lowest LOOIC (see Table XXX), but that would not always be the case.

To address the second half of the comment, we changed the candidate set of models as alluded to in response to Comment R1.3. In the original manuscript, we tested individual-level and phylogenetic effects of the composite variable \fgmax{} on a logit-scale. We realized we could achieve the same goals by modeling the effects of component traits (\gi{} and \gmax) on kinetic traits directly. We reasoned that if \fgmax{} *per se* effects kinetics, then both \gi{} and \gmax{} should be retained in candidate models and have significant effects on kinetic traits. Conversely, if only of the component traits effects kinetics, then it should be retained in the selected model, while the other is dropped. Our new approach is described in the "Model selection and parameter estimation of individual-level effects" section.

One minor discrepancy between the former and current approaches is that the current approach models the \fgmax{} on a log rather than logit-scale, since $\log \left( f_\mathrm{gmax} \right) = \log \left( g_\mathrm{i} \right) - \log \left( g_\mathrm{max} \right)$. However, the two transformations of \fgmax{} are highly correlated in our data:

```{r, echo = FALSE}

dat = read_rds("../objects/selected_model.rds")$data |>
  mutate(fgmax = exp(loggi - loggmax))

ggplot(dat, aes(qlogis(fgmax), log(fgmax))) +
  geom_point() +
  labs(
    x = expression(italic(f)[gmax]~(logit-scale)),
    y = expression(italic(f)[gmax]~(log-scale))
  ) +
  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed") +
  coord_equal() +
  theme_cowplot()

```

The new modeling approach changed our interpretation somewhat and these changes are reflected in the revised Results and Discussion <!-- TODO: cite specific instances -->. Overall, individual-level variation in \gi{} explained much more of the variation in kinetics than \gmax{}. We therefore revised the figures and mediation analyses to emphasize the role of \gi{} rather than \fgmax{}. We also note that the second best model retained \gmax, the estimated coefficient was in the predicted direction, but it was not significant.

In response to the request to present focal estimates across all plausible models, @fig-estimates-b and @fig-estimates-phy show individual-level fixed effects and phylogenetic correlations, respectively, of \gcl, \gi, and \gmax{} on both $\tau$ and $\lambda$ across all `r num2word(sum(df_comparison$plausible))` plausible models identified by LOOIC (Figures SXXX-SXXX in the Supporting Information). Estimates are qualitatively consistent across the plausible model set, supporting our decision to base inference on the selected model while also discussing alternative interpretations from the other plausible models where relevant.

```{r}
#| label: fig-estimates-b
#| echo: false
#| message: false
#| warning: false
#| fig-cap: !expr paste0("Individual-level (fixed-effect) estimates of guard cell length (\\gcl), initial stomatal conductance (\\gi), and anatomical maximum stomatal conductance (\\gmax) on the stomatal closure time-constant ($\\tau$) and lag-time ($\\lambda$), across all ", num2word(sum(df_comparison$plausible)), " plausible models identified by LOOIC (@tbl-comparison in the manuscript). Points and horizontal bars are posterior medians and 95\\% credible intervals; the selected model is marked with an asterisk (*). Black indicates estimates whose 95\\% credible interval excludes zero; grey indicates intervals that overlap zero.")

knitr::include_graphics("../figures/estimates-b.pdf")
```

```{r}
#| label: fig-estimates-phy
#| echo: false
#| message: false
#| warning: false
#| fig-cap: !expr paste0("Phylogenetic correlations between guard cell length (\\gcl), initial stomatal conductance (\\gi), and anatomical maximum stomatal conductance (\\gmax) and the stomatal closure time-constant ($\\tau$) and lag-time ($\\lambda$), across all ", num2word(sum(df_comparison$plausible)), " plausible models identified by LOOIC (@tbl-comparison in the manuscript). Points and horizontal bars are posterior medians and 95\\% credible intervals of the phylogenetic correlation; the selected model is marked with an asterisk (*). Black indicates estimates whose 95\\% credible interval excludes zero; grey indicates intervals that overlap zero.")

knitr::include_graphics("../figures/estimates-phy.pdf")
```

Finally, we report Pareto $\hat{k}$, PSIS effective sample size (ESS), and Monte Carlo standard error (MCSE) diagnostics for the LOOIC comparison, as the reviewer requested (`r/39_get-loo-diagnostics.R`; Table XXX in the Supporting Information). Across the `r num2word(sum(df_comparison$plausible))` plausible models, the vast majority of the `r unique(tbl_loo_diagnostics$n_curves)` curves had Pareto $\hat{k} \le 0.7$ (`r formatC(min(tbl_loo_diagnostics$pct_good), format = 'f', digits = 1)`–`r formatC(max(tbl_loo_diagnostics$pct_good), format = 'f', digits = 1)`% of curves), with the remainder in the "bad" (`r formatC(min(tbl_loo_diagnostics$pct_bad), format = 'f', digits = 1)`–`r formatC(max(tbl_loo_diagnostics$pct_bad), format = 'f', digits = 1)`%) rather than "very bad" range; no model had any curve with $\hat{k} \ge 1$. The minimum PSIS ESS across curves was as low as `r round(min(tbl_loo_diagnostics$min_psis_ess))` for some models, which is well below the conventional rule-of-thumb of a few hundred effective draws, so the corresponding PSIS-LOO estimates for those individual curves (and the resulting MCSE for the model as a whole) should be treated with some caution rather than taken at face value. We have added this caveat to the Materials and Methods (lines XXX).

```{r}
#| label: tbl-loo-diagnostics
#| echo: false
#| message: false
#| warning: false

tbl_loo_diagnostics |>
  mutate(
    model = str_replace(model, "_", "\\\\_"),
    model = ifelse(selected, paste0("\\textbf{", model, "}"), model),
    across(
      c(pct_good, pct_bad, pct_very_bad),
      \(.x) formatC(.x, format = "f", digits = 2)
    ),
    mcse_elpd_loo = ifelse(
      is.na(mcse_elpd_loo),
      "--",
      formatC(mcse_elpd_loo, format = "f", digits = 2)
    ),
    min_psis_ess = round(min_psis_ess)
  ) |>
  select(-selected) |>
  kbl(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    col.names = c(
      "Model", "$\\Delta$LOOIC", "SE", "$n$ curves",
      "Good $\\hat{k}$ (\\%)", "Bad $\\hat{k}$ (\\%)", "Very bad $\\hat{k}$ (\\%)",
      "Min. PSIS ESS", "MCSE(elpd\\textsubscript{loo})"
    ),
    align = c("l", rep("c", 8)),
    caption = paste0(": Pareto $\\hat{k}$, PSIS effective sample size (ESS), and Monte Carlo standard error (MCSE) diagnostics for the ", num2word(sum(df_comparison$plausible)), " plausible models identified by LOOIC (@tbl-comparison in the manuscript). The selected model is in bold. Good/bad/very bad refer to the conventional Pareto $\\hat{k}$ thresholds of 0.7 and 1. MCSE(elpd\\textsubscript{loo}) is left blank ('--') for models with at least one curve with $\\hat{k} \\ge 1$, for which the MCSE estimate is undefined.")
  ) |>
  kable_styling(latex_options = c("scale_down"))
```

:::

---

## Comment R1.8 — Single-chain Bayesian diagnostics for curve-level fits

::: revcomment
The curve-level Bayesian diagnostics are not yet convincing enough for a
dataset built from 2,124 fitted nonlinear curves. Lines 285–289 state that
each curve was fit using 1,000 post-warmup draws from a single chain, with
Rhat $< 1.05$, bulk ESS $> 400$ and fewer than 10 divergent transitions. The
code confirms the single-chain curve fitting in `r/functions.R`, lines 8–18.
I am not fully convinced with this as the only diagnostic for thousands of
nonlinear curve fits especially because \gi, $g_f$, $\tau$, and $\lambda$
are likely correlated. Stan's diagnostics treat divergent transitions as a
warning that HMC may not be fully exploring the posterior and the usual
workflow relies on multiple chains to assess convergence. Perhaps no need for
the authors to refit all 2,124 curves with four chains if that is
computationally prohibitive, but it would be great for a stratified four-chain
validation subset including difficult curves, high-$\tau$ curves, high-\fgmax{}
curves, and low-$R^2$ curves. To show or supplement that the single-chain
pipeline gives essentially the same $\tau$, $\lambda$, and \gi estimates as
a more standard multi-chain fit.
:::

::: response

<!-- DONE -->

**Response:**

We refit all curves with 4 chains instead of 1 and updated the Materials and Methods accordingly. As far as we can tell, this did not meaningfully impact any downstream analyses.

:::

---

## Comment R1.9 — Undocumented guard cell length outlier exclusions

::: revcomment
I found in `r/01_join-data.R`, three guard cell length outlier individuals are
removed with a comment referring to `r/scratch - gcl outliers.R`. I do not see
this exclusion described in the manuscript and the referenced scratch file is
not included in the package. Because guard cell length is one of the two
central anatomical predictors, it would be great to see this reported. The
same script also uses a first-available fallback for stomatal anatomical data
when the LICOR and stomatal leaflet do not uniquely match
(`r/01_join-data.R`). That may be reasonable but it could be better to
document because it might add measurement error to \gmax{}, guard cell length
and therefore \fgmax{}.
:::

::: response

<!-- TODO:
- put in correct Figure numbers for main text
-->

**Response:**

We unintentionally neglected to explain why these three individuals were considered outliers. In the Materials and Methods (lines XXX) we clarified that:

> "We excluded three individuals with extreme outlying guard cell length values. Preliminary analysis of variation in guard cell length among growth treatments, leaf types, and accessions identified these outliers based on a residual value greater than 0.45, which visually appeared as a clear break in the residual distribution."

Regarding the leaflet-matching fallback: when a leaflet was measured for stomatal anatomy but did not match the leaflet used for the LI-6800 humidity-response curve, we used the first available anatomical measurement from a different leaflet on the same leaf as a proxy, rather than discarding the individual. This fallback was triggered for `r gcl_switch_summary$n_switched` of `r gcl_switch_summary$n_total` individual-by-leaf-type combinations (`r formatC(100 * gcl_switch_summary$n_switched / gcl_switch_summary$n_total, format = 'f', digits = 2)`%). We agree this could add measurement error to \gmax{}, guard cell length, and \fgmax{} for those individuals, and we have added this frequency and a caveat to the Materials and Methods.

:::

---

## Comment R1.10 — Figure 1D color labeling appears reversed

::: revcomment
In Figure 1D, the orange point is the low \fgmax{} point on the steep part of
the curve and the blue point is the high \fgmax{} point on the shallow part of
the curve? The figure labels themselves appear correct. However, the Figure 1
caption says the larger change occurs at low \fgmax{} "blue; steep tangent
slope" and high \fgmax{} is "orange; shallow tangent slope." I found that is
reversed relative to the plotted colors?
:::

::: response

<!-- DONE -->

**Response:**

The reviewer is correct. The in-figure labels and plotted colors in Figure 1d were always consistent (low \fgmax{} = orange/steep slope, high \fgmax{} = blue/shallow slope), but we had the colors swapped in the Figure 1 caption text. We corrected the caption to match the figure.
:::

---

# Reviewer 2

## Comment R2.1 — Abstract sentence unclear

::: revcomment
I am not sure this sentence reads clearly: "Leaves with small guard cells and
a lower \gop{} to \gmax{} ratio (\fgmax{}) closed faster, but explained
variation in kinetic parameters at different levels of biological
organization." Perhaps it is missing 'both'? "Leaves with small guard cells
and a lower \gop{} to \gmax{} ratio (\fgmax{}) closed faster, but both
explained variation in kinetic parameters at different levels of biological
organization."
:::

::: response

<!-- DONE -->

**Response:**

We revised this sentence to be clearer and updated it based on new results:

> Leaves with small guard cells and a lower \gop{} closed faster, but each explained variation in kinetic parameters at different levels of biological organization.

:::

---

## Comment R2.2 — Mechanistic follow-up on turgor-responsive window

::: revcomment
This is really interesting and links back to early work on turgor maintenance
and speed: "Smaller stomata can be speedier, but only if stomata are held at
an aperture where they are responsive to changing turgor pressure." How could
the authors mechanistically follow this up? Using anatomical measurements of
$g_\mathrm{smax}$ as a benchmark for comparing to operational \gsw{} has some
drawbacks — e.g.\ over-estimation. I think some of this is covered in the
discussion.
:::

::: response
**Response:**

### MARKER - continue here

[YOUR RESPONSE HERE]
:::

---

## Comment R2.3 — Wrong-way response magnitude and \fgmax{}

::: revcomment
The authors have chosen species with 'wrong way response' to VPD — do they
take into account or explore the magnitude of the initial rise in \gsw{} and
how this might relate to lag time or the \fgmax{} relationship (the magnitude
of the 'hump' in Figure 1a)?
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R2.4 — Compensatory aperture in adaxial-blocked leaves (L403)

::: revcomment
L403 — stomatal aperture was more open in leaves where gas exchange through
adaxial stomata was blocked — is this interesting, do the authors think it is
compensatory?
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

---

## Comment R2.5 — Figure 1B legend format

::: revcomment
Fig 1b — Personal preference; although I appreciate the legend is correct, it
might be clearer to make the large and small time constants a rectangle of the
appropriate colour or dot of the colour? When I look at the figure I focus in
on the format of the line rather than the colour which distracts from
understanding the figure quickly.
:::

::: response
**Response:**

[YOUR RESPONSE HERE]
:::

# References

::: {#refs}
:::
